---
title: "biplotEZ"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{biplotEZ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(biplotEZ)
```

The package \pkg{biplotEZ} provides users with an EZ-to-use way of constructing multi-dimensional scatterplots of their data. The simplest form of a biplot is the principal component analysis (PCA) biplot which will be used for illustration in this vignette.

## What is a PCA biplot
Consider a data matrix $\mathbf{X}:n \times p$ containing data on $n$ objects and $p$ variables. To produce a 2D biplot, we need to optimally approximate $\mathbf{X}$ (typically of rank $p$ with $p<n$) with a rank $2$ matrix. In terms of the least squares error, we want to

$$
min \| \hat{\mathbf{X}}-\mathbf{X} \|
$$
where $rank(\hat{\mathbf{X}})=2$. It was shown by @EckartYoung1936 that if the singular value decomposition of $\mathbf{X} = \mathbf{UDV'}$ then
$$
\hat{\mathbf{X}} = \mathbf{UJDJV'}
$$
with 
$$
\mathbf{J} = \begin{bmatrix}
              \mathbf{I}_2 & \mathbf{0}\\
              \mathbf{0} & \mathbf{0}
             \end{bmatrix}
$$
essentially selecting only the first two columns of $\mathbf{U}$, the diagonal matrix of the first (largest) two singular values and the first two rows of $\mathbf{V}'$.


## References
