---
title: "Biplots in 3D"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Biplots in 3D}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.height = 6, fig.width = 7,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(biplotEZ)
library(rgl)
```


# PCA 3d biplot
Three-dimensional PCA biplots are obtained by specifying `dim.biplot = 3` in the call to `PCA()`. The package `rgl` is required and on calling the `plot()` the biplot is drawn in an `rgl` window. The plot can then be interactively rotated and zoomed using the mouse buttons.  


```{r }
biplot(iris) |> PCA(group.aes = iris$Species,dim.biplot = 3) |>
  axes(col="black") |> plot()
```


```{r echo=FALSE}
my.plot <- scene3d()
rglwidget(my.plot)
```


# CVA 3d biplot
Similarly for a CVA 3d biplot, `dim.biplot = 3` is specified in the call to `CVA()`. 
```{r}
biplot(iris) |> CVA(classes = iris$Species,dim.biplot = 3) |>
  axes(col="black") |> plot()
```

```{r echo=FALSE}
my.plot <- scene3d()
rglwidget(my.plot)
```


# Adding new samples
The process of adding new samples to the biplot, called interpolation utilises the functions `interpolate()` and `newsamples()`. These functions work in the same way as in the call to the two-dimensional biplot. The function `interpolate()` accepts the argument `newdata` to specify a matrix or data frame containing the new samples to be interpolated. The function `newsamples()` operates the same way as `samples()` in that the user can specify the aesthetics of the interpolated samples. 

```{r}
biplot(iris[1:100,1:4])|> PCA(dim.biplot = 3) |> axes(col="black") |> 
  interpolate(newdata = iris[101:150,1:4]) |> newsamples(col="purple") |> plot()
```

```{r echo=FALSE}
my.plot <- scene3d()
rglwidget(my.plot)
```


# Adding new variables
To interpolate new variables to the biplot, the function `interpolate()` and `newaxes()` are called. The function `interpolate()` accepts the argument `newvariable` to specify a matrix or data frame of the same number of rows in the data specified in `biplot()` containing the new variables to be interpolated. The function `newaxes()` allows the user to specify the aesthetics of the interpolated variables. 

```{r}
biplot(iris[,1:3])|> PCA(dim.biplot = 3) |> axes(col="black") |> 
  interpolate(newvariable = iris[,4]) |> 
  newaxes(col="darkred",X.new.names = "Petal.Width") |> plot()
```


```{r echo=FALSE}
my.plot <- scene3d()
rglwidget(my.plot)
```


# Predicting samples
To add the prediction of samples on the biplot, the `prediction()` function is used. The `predict.samples` argument takes in a vector indicating either the row numbers of the samples to predict or set to TRUE indicating to predict all samples. The argument `which` allows the user to select which variable to predict the sample on. In the example below, samples 100 to 150 predictions are shown for variables 1 and 4. The aesthetics for the display of the predictions are arguments in the `axes()` function: `predict.col` and `predict.lwd`. 
```{r}
biplot(iris) |> PCA(group.aes = iris$Species,dim.biplot = 3,show.class.means = TRUE) |> 
  axes(col="black",predict.col = "orange") |>  
  prediction(predict.samples=100:150,which = c(1,4))  |> plot()
```

```{r echo=FALSE}
my.plot <- scene3d()
rglwidget(my.plot)
```


# Predicting group means 
Similarly, to add the prediction of group means, the function `prediction()` is used. The argument `predict.means` takes in a vector specifying which group means to predict. In the example below, only the first group means is predicted. Important to note that the argument `show.class.means` must be set to TRUE in the `PCA()` function. 
```{r}
biplot(iris) |> 
  PCA(group.aes = iris$Species,dim.biplot = 3,show.class.means = TRUE) |> 
  axes(col="black",predict.col = "darkred") |> 
  prediction(predict.means = 1) |> plot()
```

```{r echo=FALSE}
my.plot <- scene3d()
rglwidget(my.plot)
```


# Ellipses 
Ellipses are added to a 3d biplot using the `ellipses()` function which works in the same way as a 2d biplot. 
```{r}
biplot(iris) |> PCA(group.aes = iris[,5],dim.biplot = 3) |> 
  axes(col="black") |> 
  ellipses(kappa = 3,opacity = 0.5) |> plot()
```

```{r echo=FALSE}
my.plot <- scene3d()
rglwidget(my.plot)
```
