
---
title: "Biplots in 3D"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Biplots in 3D}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.height = 6, fig.width = 7,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(biplotEZ)
library(rgl)
```

# PCA 3D biplot
Three-dimensional PCA biplots are obtained by specifying `dim.biplot = 3` in the call to `PCA()`. The package `rgl` is required and on calling the `plot()` the biplot is drawn in an `rgl` window. The plot can then be interactively rotated and zoomed using the mouse buttons.  

```{r }
biplot(iris) |> PCA(group.aes = iris$Species,dim.biplot = 3) |>
  axes(col="black") |> plot()
```

```{r echo=FALSE}
my.plot <- scene3d()
rglwidget(my.plot)
```

## Adding new samples
The process of adding new samples to the biplot, called interpolation utilises the functions `interpolate()` and `newsamples()`. These functions work in the same way as in the call to the two-dimensional biplot. The function `interpolate()` accepts the argument `newdata` to specify a matrix or data frame containing the new samples to be interpolated. The function `newsamples()` operates the same way as `samples()` in that the user can specify the aesthetics of the interpolated samples. 

```{r}
biplot(iris[1:100,1:4])|> PCA(dim.biplot = 3) |> axes(col="black") |> 
  interpolate(newdata = iris[101:150,1:4]) |> newsamples(col="purple") |> plot()
```

```{r echo=FALSE}
my.plot <- scene3d()
rglwidget(my.plot)
```

## Adding new variables
To interpolate new variables to the biplot, the function `interpolate()` and `newaxes()` are called. The function `interpolate()` accepts the argument `newvariable` to specify a matrix or data frame of the same number of rows in the data specified in `biplot()` containing the new variables to be interpolated. The function `newaxes()` allows the user to specify the aesthetics of the interpolated variables. 

```{r}
biplot(iris[,1:3])|> PCA(dim.biplot = 3) |> axes(col="black") |> 
  interpolate(newvariable = iris[,4]) |> 
  newaxes(col="darkred",X.new.names = "Petal.Width") |> plot()
```


```{r echo=FALSE}
my.plot <- scene3d()
rglwidget(my.plot)
```

## Predicting samples
To add the prediction of samples on the biplot, the `prediction()` function is used. The `predict.samples` argument takes in a vector indicating either the row numbers of the samples to predict or set to TRUE indicating to predict all samples. The argument `which` allows the user to select which variable to predict the sample on. In the example below, samples 100 to 150 predictions are shown for variables 1 and 4. The aesthetics for the display of the predictions are arguments in the `axes()` function: `predict.col` and `predict.lwd`. 
```{r}
biplot(iris) |> PCA(group.aes = iris$Species,dim.biplot = 3,show.class.means = TRUE) |> 
  axes(col="black",predict.col = "orange") |>  
  prediction(predict.samples=100:150,which = c(1,4))  |> plot()
```

```{r echo=FALSE}
my.plot <- scene3d()
rglwidget(my.plot)
```

## Predicting group means 
Similarly, to add the prediction of group means, the function `prediction()` is used. The argument `predict.means` takes in a vector specifying which group means to predict. In the example below, only the first group means is predicted. Important to note that the argument `show.class.means` must be set to TRUE in the `PCA()` function. 
```{r}
biplot(iris) |> 
  PCA(group.aes = iris$Species,dim.biplot = 3,show.class.means = TRUE) |> 
  axes(col="black",predict.col = "darkred") |> 
  prediction(predict.means = 1) |> plot()
```

```{r echo=FALSE}
my.plot <- scene3d()
rglwidget(my.plot)
```

## Ellipses 
Ellipses are added to a 3D biplot using the `ellipses()` function which works in the same way as a 2D biplot. 
```{r}
biplot(iris) |> PCA(group.aes = iris[,5],dim.biplot = 3) |> 
  axes(col="black") |> 
  ellipses(kappa = 3,opacity = 0.5) |> plot()
```

```{r echo=FALSE}
my.plot <- scene3d()
rglwidget(my.plot)
```

# CVA 3d biplot
Similarly for a CVA 3D biplot, `dim.biplot = 3` is specified in the call to `CVA()`. 
```{r}
biplot(iris) |> CVA(classes = iris$Species,dim.biplot = 3) |>
  axes(col="black") |> plot()
```

```{r echo=FALSE}
my.plot <- scene3d()
rglwidget(my.plot)
```

# CA 3D biplot

As with `PCA()`, three-dimensional CA biplots are obtained by specifying `dim.biplot = 3` in the call to `CA()`. Again, an interactive biplot will be drawn in a separate `rgl` window.

Consider the `HairEyeColor` example again as discussed in `CA in biplotEZ`:

```{r}
biplot(HairEyeColor[,,2], center = FALSE) |> 
  CA(variant = "Symmetric", dim.biplot = 3) |>
  samples(col=c("darkred","forestgreen"), pch=c(15,17), label.col="black") |>
  plot()
```

```{r echo=FALSE}
my.plot <- scene3d()
rglwidget(my.plot)
```

We also illustrate the three dimensional CA biplot of the `SA Crime` example: 

```{r}
SACrime <- matrix(c(1235,432,1824,1322,573,588,624,169,629,34479,16833,46993,30606,13670,
              16849,15861,9898,24915,2160,939,5257,4946,722,1271,881,775,1844,5946,
              4418,15117,10258,5401,4273,4987,1956,10639,29508,15705,62703,37203,
              11857,18855,14722,4924,42376,604,156,7466,3889,203,664,291,5,923,19875,
              19885,57153,29410,11024,12202,10406,5431,32663,7086,4193,22152,9264,3760,
              4752,3863,1337,8578,7929,4525,12348,24174,3198,1770,7004,2201,45985,764,
              427,1501,1197,215,251,345,213,1850,3515,879,3674,4713,696,835,917,422,2836,
              88,59,174,76,31,61,117,32,257,5499,2628,8073,6502,2816,2635,3017,1020,4000,
              8939,4501,50970,24290,2447,5907,5528,1175,14555),nrow=9, ncol=14)
dimnames(SACrime) <- list(paste(c("ECpe", "FrSt", "Gaut", "KZN",  "Limp", "Mpml", "NWst", "NCpe",
                            "WCpe")), paste(c("Arsn", "AGBH", "AtMr", "BNRs", "BRs",  "CrJk",
                                              "CmAs", "CmRb", "DrgR", "InAs", "Mrd", "PubV", 
                                              "Rape", "RAC" )))
names(dimnames(SACrime))[[1]] <- "Provinces"
names(dimnames(SACrime))[[2]] <- "Crimes"
```

```{r}
biplot(SACrime, center = FALSE) |> CA(variant = "Symmetric", dim.biplot = 3) |> 
samples(col=c("royalblue","darkred"), pch=c(15,17), label.col="black") |> plot()
```

```{r echo=FALSE}
my.plot <- scene3d()
rglwidget(my.plot)
```