---
title: "CA in biplotEZ"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{CA in biplotEZ}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.height = 6, fig.width = 7,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(biplotEZ)
```

# What is a Correspondence Analysis?

In its simplest form Correspondence Analysis (CA) aims to expose the
association between two categorical variables by utilising a two-way
frequency table. Numerous variants of CA are available for the
application to diverse problems, the interested reader is referred to:
@UB2011, @BehLombardo2014.

In the <a href="https://cran.rstudio.com/web/packages/biplotEZ/vignettes/biplotEZ.html" target="_blank">biplotEZ vignette</a> focus will be places on three *EZ*-to-use versions based on the Pearson residuals (@UB2011 :300).

Now, the two-way frequency table is also referred to as the data matrix:
$\mathbf{X}:r\times c$. This data matrix is different from the
continuous case used for the `PCA()` and `CVA()` examples, as it
represents the cross-tabulations of two categorical variables (i.e.
factors), each with a finite number of levels (i.e response values). The
elements of the data matrix represent the frequency of the co-occurrence
of two particular levels of the two variables. Consider the
`HairEyeColor` data set in `R`, which summarises the hair and eye color
of male and female statistics students. For the purpose of this example
only the male students will be considered:

```{r}
X <- HairEyeColor[,,2]
X
```

The grand total of the table $N$ is obtained from the total of all
frequencies:

$$
\sum_{r=1}^{R}\sum_{c=1}^{C}x_{rc}=N
$$
```{r}
N <- sum(X)
N
```

It is common to work with the proportions rather than the frequencies
in terms of the correspondence matrix, $\mathbf{P}$:

$$
\mathbf{P}=\frac{\mathbf{X}}{N}
$$ 
```{r}
P <- X/N
P
```

Other useful summaries of $\mathbf{P}$ include the row and column masses (for arbitrary row and column $r$ and $c$, respectively), also expressed as diagonal matrices:

$$
\mathbf{r}_r = \sum_{c=1}^{C}p_{rc}; \hspace{0.5 cm}
\mathbf{c}_c = \sum_{r=1}^{R}p_{rc}\\
\mathbf{r}=\mathbf{P1}; \hspace{0.5 cm} \mathbf{c}=\mathbf{P}^\prime\mathbf{1}
$$
```{r}
rMass <- rowSums(P)
rMass
cMass <- colSums(P)
cMass
```

Diagonal matrices:

$$
\mathbf{D_r}=\text{diag}(\mathbf{r}); \hspace{0.5 cm} \mathbf{D_c}=\text{diag}(\mathbf{c})
$$
```{r}
Dr <- diag(apply(P, 1, sum))
Dr
Dc <- diag(apply(P, 2, sum))
Dc
```
In order to obtain the first form of the row and column coordinates, the singular value decomposition (SVD) of the matrix of standardised Pearson residuals ($\mathbf{S}$) is computed:

$$
\begin{aligned}
\text{SVD}(\mathbf{S}) &= \text{SVD}\left(\mathbf{D_r^{-\frac{1}{2}}}(\mathbf{P}-\mathbf{rc^\prime})\mathbf{D_c^{-\frac{1}{2}}}\right)\\&= \mathbf{U\Lambda V^\prime}
\end{aligned}
$$
The return value for the Standardised pearson residuals is `Smat` and the singular value decomposition, `SVD`.

```{r}
Smat <- sqrt(solve(Dr))%*%(P-(Dr %*%matrix(1, nrow = nrow(X), 
    ncol = ncol(X)) %*%  Dc))%*%sqrt(solve(Dc))
svd.out <- svd(Smat)
svd.out
```
This is linked to the $\chi^2$-statistic to determine whether the two categorical variables (i.e. the rows and columns of the contingency table) are independent. The expected frequencies represented by the product of the row and column masses ($\mathbf{rc^\prime}$). 

Furthermore, since the weights of certain objects might be substantially different from others which could result in a distorted approximation in lower dimension, the $\chi^2$-distance, also referred to as the weighted Euclidean distance, is rather used to measure distances in CA.
This is an intuitive decision as it follows from the $\chi^2$-statistic to test the independence between two categorical variables, in this case the independence between the rows and columns of the contingency table. (@BehLombardo2014, @Greenacre2017).

# CA biplot

## Coordinates

In order to construct a biplot in which the distances between the row and column coordinates are meaningful an asymmetric display should be constructed. This means that the contribution of the singular values should be different for the row and column coordinates. (@Gabriel1971)
The standard coordinates are expressed by:

$$
\begin{aligned}
\text{Row standard coordinates:} \hspace{0.5 cm}&\mathbf{U}\\
\text{Column standard coordinates:} \hspace{0.5 cm}&\mathbf{V}
\end{aligned}
$$

The principal coordinates are expressed by:

$$
\begin{aligned}
\text{Row principal coordinates:} \hspace{0.5 cm}&\mathbf{U\Lambda}\\
\text{Column principal coordinates:} \hspace{0.5 cm}&\mathbf{V\Lambda}
\end{aligned}
$$

By including the singular values the magnitude of the association between the variables are incorporated in the scaling of the coordinates.

In the `ca()` function the argument `variant` allows the user to choose between three types of CA biplots: `Princ`, `Stand` and `Symmetric`.

$$
\begin{aligned}
\text{Row coordinates:} \hspace{0.5 cm}&\mathbf{U\Lambda^\gamma}\\
\text{Column coordinates:} \hspace{0.5 cm}&\mathbf{V\Lambda^{1-\gamma}}
\end{aligned}
$$
The row standard (i.e. column principal) coordinate biplot: `Stand`, results from $\gamma=0$.

The row principal (i.e. column standard) coordinate biplot: `Princ`, results from $\gamma=1$.

The symmetric plot in which row and column coordinates are scaled equally: `Symmetric`, results from $\gamma=0.5$.

The return value is `rowcoor` and `colcoor`, respectively.

```{r}
gamma <- 0
rowcoor <- svd.out[[2]]%*%(diag(svd.out[[1]])^gamma)
colcoor <- svd.out[[3]]%*%(diag(svd.out[[1]])^(1-gamma))
```

## Lambda scaling

As presented in @UB2011 :24, when constructing a biplot representing the rows of a coordinate matrix $\mathbf{A}$ and $\mathbf{B}^\prime$. Take note that the inner product is invariant when $\mathbf{A}$ and $\mathbf{B}$ are scaled inversely by $\lambda$. 

$$
\mathbf{AB} = (\lambda\mathbf{A})(\mathbf{B}/\lambda)
$$
An arbitrary value of $\lambda$ can be selected or an *optimal* value could be to ensure that the average squared distance of the points in $\lambda\mathbf{A}$ and $\mathbf{B}/\lambda$ is equal.

$$
\lambda^4 =\frac{r}{c}\frac{||\mathbf{B}||^2}{||\mathbf{A}||^2}
$$

The default setting is to not apply lambda-scaling (i.e. $\lambda=1$).

The return value is `lambda.val`.

## Measures of fit

`quality`, `adequacy`, `row.predictivities` and `column.predictivities` are available for CA biplots. 

As explained in the <a href="https://cran.rstudio.com/web/packages/biplotEZ/vignettes/biplotEZ.html" target="_blank">biplotEZ vignette</a>, the quality of the biplot is measured by the ratio of the variance explained (sum of the squared singular values of the utilised ($M$) components) and the total variance (sum of all squared singular values ($p$)).

$$
\frac{\sum_{m=1}^{M}\lambda_m^2}{\sum_{m=1}^{p}\lambda_m^2}
$$
The `adequacy` refers to the representation of the variables. In `CA()` the factor variable represented in the columns is treated as the variables and is calculated as explained in the <a href="https://cran.rstudio.com/web/packages/biplotEZ/vignettes/biplotEZ.html" target="_blank">biplotEZ vignette</a>:

$$
 \frac{diag(\mathbf{V}_r\mathbf{V}_r')}{diag(\mathbf{VV}')}= diag(\mathbf{V}_r\mathbf{V}_r')
$$


The predictivities provide a measure of how well the original values are recovered from the biplot. An element that is well represented will have a predictivity close to one, indicating that the row or column variable values from prediction is close to the observed values. If an element is poorly represented, the predicted values will be very different from the original values and the predictivity value will be close to zero.

The `row.predictivities` are calculated as follows (@UB2011 :299):

$$
diag(\mathbf{U}\mathbf{\Sigma}\mathbf{J}\mathbf{\Sigma}\mathbf{U}')[diag(\mathbf{U}\mathbf{\Sigma}\mathbf{\Sigma}\mathbf{U}')]^{-1}\\
=diag(\mathbf{U}\mathbf{\Sigma}^2\mathbf{J}\mathbf{U}')[diag(\mathbf{U}\mathbf{\Sigma^2}\mathbf{U}')]^{-1}
$$

The `col.predictivities` are calculated as follows (@UB2011 :299):

$$
diag(\mathbf{V}\mathbf{\Sigma}^2\mathbf{J}\mathbf{V}')[diag(\mathbf{V}\mathbf{\Sigma^2}\mathbf{V}')]^{-1}
$$

# The function CA()

The function `CA()` requires a two-way contingency table as input and will return an object of class `CA` and `biplot`. As this is not a standard data matrix as for `PCA` and `CVA`, scaling and centering is not allowed on the two-way contingency table and a warning will be given if either `scale` or `center` is specified as `TRUE` in biplot()`.

## `Variant="Princ"`

The default CA biplot is a row principal coordinate biplot:

```{r}
biplot(HairEyeColor[,,2], center = FALSE) |> CA() |> plot()
```

## `Variant="Stand"`

To construct the CA biplot for row standard coordinates:

```{r}
ca.out <- biplot(HairEyeColor[,,2], center = FALSE) |> CA(variant = "Stand") |> 
  plot()
```

## `Variant="Symmetric"`

To construct the symmetric CA map:

```{r}
ca.out <- biplot(HairEyeColor[,,2], center = FALSE) |> 
  CA(variant = "Symmetric") |> plot()
ca.out$lambda.val
```


## Measures of fit

The `fit.mesaures()` function should be utilised to obtain the specific fit measures explained above.

```{r}
ca.out <- biplot(HairEyeColor[,,2], center = FALSE) |> 
  CA(variant = "Symmetric") |> fit.measures()
print("Quality")
ca.out$quality
print("Adequacy")
ca.out$adequacy
print("Row predictivities")
ca.out$row.predictivities
print("Column predictivities")
ca.out$col.predictivities
```

## Interpolating new samples

Adding cross-tabulations of the two categorical variables to the plot is facilitated by the function `interpolate()`. Note that the additional variables to be interpolated did not contribute to the construction of the biplot. This is the reason why @Greenacre2017 term these supplementary points. 

The function `interpolate()` accepts a matrix or data frame containing the samples and variables to be interpolated. The argument `newdata` containing the samples to be interpolated needs to have a similar structure to the data set sent to `biplot()`. If `biplot()` received a data frame, `newdata` can be either another data frame or a matrix containing the subset of numerical variables.


The function `newsamples()` operates similar to `samples()` and enables aesthetic changes to the new samples.

```{r}
biplot(HairEyeColor[,,2], center = FALSE) |> CA(variant = "Symmetric") |> 
  samples(pch = c(0,2)) |> interpolate(newdata = HairEyeColor[,,1]) |> 
  newsamples(col = c("orange","purple"), pch = c(15,17)) |> plot()      
```


## Aesthetics and legend

The `sample()` function should be utilised to specify the colours, plotting characters and expansion of the samples.

```{r}
biplot(HairEyeColor[,,2], center = FALSE) |> CA(variant = "Princ") |> 
  samples(col = c("cyan","purple"), pch = c(15,17), label.side = c("bottom","top"), 
          label.cex = 1) |> legend.type(samples = TRUE, new = TRUE) |> plot()
```

```{r}
biplot(HairEyeColor[,,2], center = FALSE) |> CA(variant = "Symmetric") |> 
  samples(col = c("forestgreen","magenta"), pch = c(12,17), 
          label.side = c("top","bottom")) |> 
  legend.type(samples = TRUE) |> plot()
```

## Additional example

Consider the South African Crime data set 2008, extracted from the South African police website (http://www.saps.gov.za/). @UB2011 :312.

```{r}
SACrime <- matrix(c(1235,432,1824,1322,573,588,624,169,629,34479,16833,46993,30606,13670,
              16849,15861,9898,24915,2160,939,5257,4946,722,1271,881,775,1844,5946,
              4418,15117,10258,5401,4273,4987,1956,10639,29508,15705,62703,37203,
              11857,18855,14722,4924,42376,604,156,7466,3889,203,664,291,5,923,19875,
              19885,57153,29410,11024,12202,10406,5431,32663,7086,4193,22152,9264,3760,
              4752,3863,1337,8578,7929,4525,12348,24174,3198,1770,7004,2201,45985,764,
              427,1501,1197,215,251,345,213,1850,3515,879,3674,4713,696,835,917,422,2836,
              88,59,174,76,31,61,117,32,257,5499,2628,8073,6502,2816,2635,3017,1020,4000,
              8939,4501,50970,24290,2447,5907,5528,1175,14555),nrow=9, ncol=14)
dimnames(SACrime) <- list(paste(c("ECpe", "FrSt", "Gaut", "KZN",  "Limp", "Mpml", "NWst", "NCpe",
                            "WCpe")), paste(c("Arsn", "AGBH", "AtMr", "BNRs", "BRs",  "CrJk",
                                              "CmAs", "CmRb", "DrgR", "InAs", "Mrd", "PubV", 
                                              "Rape", "RAC" )))
names(dimnames(SACrime))[[1]] <- "Provinces"
names(dimnames(SACrime))[[2]] <- "Crimes"
SACrime
```

```{r}
biplot(SACrime, center = FALSE) |> 
  CA(variant = "Symmetric", lambda.scal = TRUE) |> 
  samples(col = c("cyan","purple"), pch = c(15,17), label.side = c("bottom","top")) |>
  legend.type(samples = TRUE, new = TRUE) |> plot()
```


# References
